Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-15 docs/paper.Rmd
replace volume/issue template with publication month and year only

The trade-off between child quantity and child quality is a foundational concept in the economics of the family. First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{beckerlewis1973, beckertomes1976}, this framework posits that parents allocate finite resources—both financial and non-financial—across children. An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital. This mechanism has served as an explanatory model for changes in fertility behavior and the evolution of population structures in low- and middle-income countries.

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2010fertility} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

This research contributes to the literature in several respects. It introduces new evidence on the nutritional consequences of fertility shocks in an urban developing-country context. It expands the measurement of child quality beyond educational outcomes to include biological indicators. It also adds to the limited body of work that uses plausibly exogenous variation in reproductive health policy to study downstream effects on human capital. More broadly, the findings underscore the potential for local family planning restrictions to generate unintended consequences for child well-being in settings characterized by poverty, food insecurity, and institutional fragility.

>>>>

The trade-off between child quantity and child quality is a foundational concept in the economics of the family. First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{becker1973interaction, becker1976child}, this framework posits that parents allocate finite resources—both financial and non-financial—across children. An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital. This mechanism has served as an explanatory model for changes in fertility behavior and the evolution of population structures in low- and middle-income countries.

@@

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

This study contributes to the literature in several important ways. It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities. Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting. It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being. More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.

\section{Review of Related Literature}

The quantity–quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each. Quality in this context refers to the human capital of each child: attributes like education, health, and nutrition that enhance a child’s future productivity and well-being. The genesis of this idea traces back to Gary Becker’s seminal work around 1960, which for the first time treated children as economic goods subject to parental choice and budget constraints \citep{becker1960economic}. Becker argued that as families become wealthier, they may not simply want more children, but rather better-raised children, much as a household might prefer a higher-quality car or house over a greater quantity of them. This proposition led to a formal theory in which increases in income or changes in economic conditions cause parents to substitute child quality for quantity, consistent with historical patterns of lower fertility and higher educational attainment during economic development \citep{galor2000population}.

In what follows, I review the theoretical foundations of the Q-Q model and its evolution in the literature. I begin with the static models of Becker \citep{becker1960economic} and Becker–Lewis \citep{becker1973interaction}, which first formalized the trade-off within a household utility maximization framework. We then examine extensions to dynamic, intergenerational settings, including the contributions of Becker and Tomes \citep{becker1976child} on child endowments and the altruistic dynastic model associated with Barro and Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and unified growth models, notably Galor and Weil \citep{galor2000population} and Galor and Moav \citep{galor2002natural}, which integrate the Q–Q mechanism into a general theory of demographic and economic transformation.

Finally, I discuss more recent refinements that enrich the basic model by incorporating credit constraints \citep{doepke2004accounting}, intra-household bargaining \citep{doepke2019bargaining}, and multi-dimensional child quality \citep{hoddinott2013economic, kalemli2002does}, with a special emphasis on health and nutrition. My focus is on how the Q–Q framework has been applied to understand fertility and child investment patterns, especially in developing country contexts where resource constraints and health outcomes are paramount.

\subsection{Theoretical Background}
